---
title: "R Notebook"
output: html_notebook
author: Abraham Lewis
Date: December 3,2018 
---


```{r}
library(readr)
library(ggplot2)
library(dplyr)
options(scipen=999)
```

This data is from kaggle and is based of the world health orginaztiions morality tool.

No complete count is kept of suicide attempts in the U.S.; however, each year the CDC (Center for diesies control prevention) gathers data from hospitals on non-fatal injuries from self-harm as well as survey data.




```{r}
 who<-read_csv("who_suicide_statistics.csv")
```




```{r}
 nrow(who)
```


```{r}
ncol(who)
```


```{r}
glimpse(who)
```




```{r}
select(who, country) %>% unique %>% nrow 
```





```{r}
unique(who$country) 
```
141

Speaking at an event on suicide prevention in capital Kabul, Afghanistan Independent Human Rights Commission (AIHRC) Commissioner Qudri Yazdan said while suicides continue to be on the rise in the country, several cases go unreported.

The first being afganistan





```{r}
 wc <-read.csv("countries of the world.csv")
```

```{r}
names(wc)
```
```{r}
unique(wc$Country)
```



KABUL, Afghanistan - Up to 3,000 people are committing suicide in Afghanistan each year, 80 percent of whom are women, an official said Sunday.




The Un's website and more importantly wikepedia says


195 Countries


54 countries are in Africa
48 in Asia
44 in Europe
33 in Latin America and the Caribbean
14 in Oceania
2 in Northern America
 No county begins with w or x or y in this data set it is missing yemen.
loooking at the data we can see that it doesnt inculde all the countries.
most of the countries missing are small counrties (such as lichtenstin or kosovo) or countries in southeast asia Laos or cambodia or in africa (such as chad or mozambque)

Most Importantly its missing india and china.
So in the first data set there a numerious countries missing the first bineg afgantant

In the second list there are countries that no longer exsist such as Netherland antillies and it includes territories which have do not have the exact un status of country or are repitons of

Or Data that Just isnt collected







```{r}
summary(who)
```










```{r}
which(!complete.cases(who))
```
so nearly 13000 Na's

```{r}
head(who)

```


```{r}
       c1<- who%>%
group_by(country) %>% 
  summarise(total_suicides = sum(suicides_no,na.rm=TRUE)) %>%
  arrange(desc(total_suicides))
```


```{r}
c1
```

From this I saw that about half the countries had a suicide total over 10000 then went down all the way to Montserrat which had 1 also another fact was that about 70 percent of the countries had at least 5 years of usable data recorded.






```{r}
whoDF_s1 <- who %>% 
  group_by(country) %>% 
  summarise(total_suicides = sum(suicides_no,na.rm=TRUE)) %>%
  arrange(desc(total_suicides))

ggplot(whoDF_s1,aes(x=reorder(country,-total_suicides),y=total_suicides,fill=-total_suicides))+
  geom_col() + theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  labs(x="Country",y="Count",title="Country-wise Suicides Stats")+
  theme(plot.title = element_text(size=15,face="bold"))

```


So with the first attept to look at the data it was a mess so I decided to just look at countries in terms of age and gender with regards to suicide.

```{r}
TC <- who %>% 
  select(country, year, sex, age, suicides_no, population,) %>%
  filter(country %in% c("Russian Federation","United States of America","Japan","France","Ukraine","Germany","Republic of Korea","Brazil","Poland","United Kingdom" ))

```



```{r}

whoDF_sa <- who %>% 
  group_by(age) %>% 
  summarise(total_suicides = sum(suicides_no,na.rm=TRUE)) %>%
  arrange(desc(total_suicides))

ggplot(whoDF_sa,aes(x=reorder(age,-total_suicides),y=total_suicides,fill=-total_suicides))+
  geom_col() + theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  labs(x="Age",y="Count",title=" Age Suicides Stats")+
  theme(plot.title = element_text(size=15,face="bold"))

```

So one of the suprising things was the age the age that had the greatest number of suicides was middle aged men.


```{r}
whoDF_sb <- who %>% 
  group_by(sex) %>% 
  summarise(total_suicides = sum(suicides_no,na.rm=TRUE)) %>%
  arrange(desc(total_suicides))

ggplot(whoDF_sb,aes(x=reorder(sex,-total_suicides),y=total_suicides,fill=-total_suicides))+
  geom_col() + theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  labs(x="Gender",y="Count",title="Suicide by Gender")+
  theme(plot.title = element_text(size=15,face="bold"))
```


```{r}
whoDF_s3 <- who %>% 
  group_by(year,country) %>% 
  summarise(total_suicides = sum(suicides_no,na.rm=TRUE)) %>%
  arrange(desc(total_suicides))

ggplot(whoDF_s3,aes(x=year,y=total_suicides,fill=-total_suicides))+
  geom_col() +
  labs(x="Year",y="Count",title="Suicides Worldwide")+
  theme(plot.title = element_text(size=15,face="bold"))
```







Then to make things clearer I looked at the top ten countries. 
 
 
```{r}
whoDF_s8 <- TC%>% 
  group_by(country) %>% 
  summarise(total_suicides = sum(suicides_no,na.rm=TRUE)) %>%
  arrange(desc(total_suicides))

ggplot(whoDF_s8,aes(x=reorder(country,-total_suicides),y=total_suicides,fill=-total_suicides))+
  geom_col() + theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  labs(x="Country",y="Count",title="Top Ten Countries")+
  theme(plot.title = element_text(size=18,face="bold"))
```


```{r}
df_top10 <- who %>% 
  filter( (country== "Russian Federation")| (country== "United States of America") | (country=="Japan")| (country=="France")|  (country== "Ukraine") |(country=="Germany") |(country=="Republic of Korea")|(country=="Brazil")|(country== "United Kingdom")| (country== "Poland")   )
```

```{r}
df7<- who%>%
filter(   (country=="Japan")| (country=="France")|  (country== "Ukraine") |(country=="Germany") |(country=="Republic of Korea")|(country=="Brazil")|(country== "United Kingdom")   )

```






```{r}
whoDF_10 <- df_top10 %>% 
  group_by(year,country) %>% 
  summarise(total_suicides = sum(suicides_no,na.rm=TRUE)) %>%
  arrange(desc(total_suicides))

ggplot(whoDF_10,mapping = aes(x=year, y=total_suicides, colour = country)) +geom_line(aes(linetype = country))

```

The weird drop in Russia and Poland was caused by Na's and zeros.



```{r}
hop <- df7 %>% 
  group_by(year,country) %>% 
  summarise(total_suicides = sum(suicides_no,na.rm=TRUE)) %>%
  arrange(desc(total_suicides))

ggplot(data = hop, mapping = aes(x =country , y = total_suicides)) + 
  geom_boxplot() +
  coord_flip()
```










While Russia and the USA had the highest sum of suicides another important way to look at this issue is with the ratio of population.


```{r}
df_country_group <- who %>%
group_by(country) %>%
summarise(sumsui= sum( as.numeric(suicides_no)),popsum = sum(as.numeric(population)))
df_country_group <- na.omit(df_country_group)
df_country_group_ratio<- df_country_group %>% mutate(ratioSP =sumsui/popsum)
df_country_group_ratio <- df_country_group_ratio %>% arrange(-ratioSP)
df_country_group_ratio_top <- head(df_country_group_ratio,10)
```

```{r}
 df_country_group_ratio_top1 <- df_country_group_ratio_top %>%
   arrange(ratioSP)
ggplot(df_country_group_ratio_top1,aes(x = ratioSP,   y = country )) +
  scale_y_discrete(limits= df_country_group_ratio_top1$country) +
 geom_segment( aes(xend=0,yend=country),size = 1,color='red2')+
  geom_point(fill="red2",color="green",size=4,shape=21,stroke=2) +
  ggtitle("Countries with highest suicide to population ratio")+
   labs(x="Ratio : Suicide/Population", y="Country")
```






```{r}
 

sum(who$suicides_no,na.rm=TRUE)
```




```{r}
whoDF_s4 <- who %>% 
  group_by(year,age) %>% 
  
  summarise(total_suicides = sum(suicides_no,na.rm=TRUE)) %>%
  arrange(desc(total_suicides))

ggplot(whoDF_s4,aes(x=year,y=total_suicides,fill=age))+
  geom_col() + theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  labs(x="Year",y="Count",title="Worldwide Suicide Rates - Year and Age-group")+
  theme(plot.title = element_text(size=15,face="bold"))
```





```{r}


```

What one sees is that there is spike in the late 90's then a slight decrease and increase around 2007-2008 then a continued decrease.
 
 also the gap in the years of 84 and 85 are coused by the russian Data


I wanted to finally compare the 2 countries with highest suicide by sum, the USA and Russia.




```{r}
 whoDF_s3 <- who %>% 
  filter(country =="United States of America") %>%
  group_by(year) %>% 
  summarise(total_suicides = sum(suicides_no,na.rm=TRUE)) %>%
  arrange(desc(total_suicides))

ggplot(whoDF_s3,aes(x=year,y=total_suicides,fill=-total_suicides))+
  geom_col() +
  labs(x="Year",y="Count",title="Suicides in USA")+
  theme(plot.title = element_text(size=15,face="bold"))


```





```{r}
whoDF_s3 <- who %>% 
  filter(country == "Russian Federation") %>%
  group_by(year) %>% 
  summarise(total_suicides = sum(suicides_no,na.rm=TRUE)) %>%
  arrange(desc(total_suicides))

ggplot(whoDF_s3,aes(x=year,y=total_suicides,fill=-total_suicides))+
  geom_col() +
  labs(x="Year",y="Count",title="Suicides in Russia")+
  theme(plot.title = element_text(size=15,face="bold"))
```



```{r}
whoDF_s4 <- who %>% 
  filter(country == "United States of America") %>%
  group_by(year,age) %>% 
  summarise(total_suicides = sum(suicides_no,na.rm=TRUE)) %>%
  arrange(desc(total_suicides))

ggplot(whoDF_s4,aes(x=year,y=total_suicides,fill=age))+
  geom_col() + theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  labs(x="Year",y="Count",title="Suicides in USA - Year and Age-group")+
  theme(plot.title = element_text(size=15,face="bold"))
```
```{r}
whoDF_s4 <- who %>% 
  filter(country == "Russian Federation") %>%
  group_by(year,age) %>% 
  summarise(total_suicides = sum(suicides_no,na.rm=TRUE)) %>%
  arrange(desc(total_suicides))

ggplot(whoDF_s4,aes(x=year,y=total_suicides,fill=age))+
  geom_col() + theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  labs(x="Year",y="Count",title="Suicides in Russia - Year and Age-group")+
  theme(plot.title = element_text(size=15,face="bold"))
```




```{r}
whoDF_s6 <- who %>% 
  filter(country == "Russian Federation",year%in%c("1997","2015","1980")) %>%
  group_by(sex,age,year) %>% 
  summarise(total_suicides = sum(suicides_no,na.rm=TRUE))

ggplot(whoDF_s6,aes(x=factor(age,levels = c("5-14 years","15-24 years","25-34 years","35-54 years","55-74 years","75+ years")),y=total_suicides,fill=sex))+
  geom_col() + theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  facet_wrap(whoDF_s6$year) +
  labs(x="Age",y="Count",title="Suicides in Russia in 2015")+
  theme(plot.title = element_text(size=15,face="bold"))
```




```{r}
whoDF_s7 <- who %>% 
  filter(country == "United States of America",year%in%c("1997","2015","1980")) %>%
  group_by(sex,age,year) %>% 
  summarise(total_suicides = sum(suicides_no,na.rm=TRUE))

ggplot(whoDF_s7,aes(x=factor(age,levels = c("5-14 years","15-24 years","25-34 years","35-54 years","55-74 years","75+ years")),y=total_suicides,fill=sex))+
  geom_col() + theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  facet_wrap(whoDF_s6$year) +
  labs(x="Age",y="Count",title="Suicides in USA")+
  theme(plot.title = element_text(size=15,face="bold"))
```


In 2015, 505,507 people visited a hospital for injuries due to self-harm. This number suggests that for every reported suicide death, approximately 11.4 people visit a hospital for self-harm related injuries. However, because of the way these data are collected, we are not able to distinguish intentional suicide attempts from non-intentional self-harm behaviors.

Based on the 2016 National Survey of Drug Use and Mental Health it is estimated that 0.5 percent of the adults aged 18 or older made at least one suicide attempt.	This translates to approximately 1.3 million adults. Adult females reported a suicide attempt 1.2 times as often as males. Further breakdown by gender and race are not available.

Based on the 2015 Youth Risk Behaviors Survey, 8.6 percent of youth in grades 9-12 reported that they had made at least one suicide attempt in the past 12 months.	Girls attempted twice as often as boys (11.6% vs. 5.5%) and teens of Hispanic origin reported the highest rate of attempt (11.3%), especially Hispanic females (15.1%) when compared with white students (6.8%) and White females (9.8%). Approximately 2.8 percent reported making a suicide attempt that required treatment by a doctor or nurse.	For those requiring treatment, rates were highest for Hispanic students with black males (4.0%) and Hispanic males (2.9%) having higher rates than white male (0.9%) students.

When it comes to suicide and suicide attempts there are rate differences depending on demographic characteristics such as age, gender, ethnicity and race. Nonetheless, suicide occurs in all demographic groups.


American foundation for for prevention of suicide



In conclusion Suicide is epidemic with a large amount of causes but a dearth of solutions.
There needs to be more reaserch conducted on the subject.



